A Spectral Analysis of the Domain Decomposed Adjoint Neumann-Ulam Method

### 2012-11-30 doc/mandc2013/notes/spectral_theory.tex
adding mandc2013 template for spectral analysis paper

  The domain decomposed behavior of the adjoint Neumann-Ulam method
  for solving linear operator equations is analyzed using the spectral
  properties of the linear operator. Relationships for the average
  length of the adjoint random walks and the fraction of histories
  leaking from a domain in the decomposition are made with respect to
  the Eigenvalues and other properties of the linear operator. In
  addition, a formal definition of the optical thickness of a domain
  in the decomposition is presented based on the spectral
  analysis. The one-speed, two-dimensional neutron diffusion equation
  is used as a model problem to test the spectral theory for symmetric
  operators. In general, the spectral theory shows good agreement with
  measured computational results.

@@

\subsection{Domain Leakage}
Finally, we seek to measure the leakage from a domain in a domain
decomposed Monte Carlo calculation and assess the quality of our
analytic relation for the optical thickness of domain and the
associated leakage approximations. For this experiment, square grid
with $h=0.1$ and non-reentrant current boundary conditions. The grid
is then decomposed into 9 square domains, 3 in each cardinal
direction. For the cross sections, the absorption cross was varied
from 1 to 100 while the scattering cross section was set to zero to
create a purely absorbing environment. The optical thickness of these
domains will vary as a function of the absorption cross section if the
other parameters are fixed. To compute the optical thickness, along
with the spectral radius as given by Eq~(\ref{eq:iteration_radius}),
we also need the parameters $n_i$ and $n_s$ which respectively
describe the typical domain length and the average number of states
moved along that typical length per history transition. For our grid
above, the domains are varied in size with $50 \times 50$, $100 \times
100$, and $200 \times 200$ cells giving $n_i=50$, $n_i=100$, and
$n_i=200$ grid points or states along the typical length of the domain
respectively. Looking at the Laplacian stencil in
Figure~\ref{fig:stencil}, we see that all history transitions will
only move a single state in either the $i$ or $j$ directions due to
the symmetry of the problem. Furthermore, if we choose the $i$
direction, not all states we will transition to will move the history
in that direction. Therefore, we look to the definition of the
iteration matrix in Eq~(\ref{eq:iteration_stencil}) and the definition
of the adjoint probability matrix in Eq~(\ref{eq:adjoint_probability})
to estimate the $n_s$ parameter. For a particular transition starting
at state $(i,j)$, 6 of the 8 possible new states in the stencil move
the history in $i$ direction with
giving $n_i = \frac{3}{5}$.

To compute the leakage fraction numerically, over the global domain
\sn{1}{5} histories were sampled from a uniform source of strength
unity. At the start of a stage of histories, the number of histories
starting in the domain was computed and as the stage progressed, the
number of histories that exited the domain was tallied with the ratio
of the two numbers providing a numerical measure for the leakage
fraction. Figure~\ref{fig:measured_length} gives the domain leakage
measurements for the domain in the center of the global grid as well
as the analytic result computed by
Eq~(\ref{eq:analytic_domain_leakage}) as a function of the absorption
cross section. Again, we note good qualitative agreement between the
measured and analytic quantities.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in,clip]{leakage_variation.png}
  \end{center}
  \caption{\textbf{Measured and analytic domain leakage as a function
      of the iteration matrix spectral radius.} \textit{To test the
      behavior with respect to domain size, $n_i=50$, $n_i=100$,and
      $n_i=200$ were used. The red data was computed numerically by a
      domain-decomposed adjoint Neumann-Ulam implementation, the black
      fine-dashed data was generated by
      Eq~(\ref{eq:mean_chord_domain_leakage}) using the mean-chord
      approximation, and the dashed-dotted black data was generated by
      Eq~(\ref{eq:wigner_domain_leakage}) using the Wigner rational
      approximation.}}
  \label{fig:measured_leakage}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in,clip]{leakage_error.png}
  \end{center}
  \caption{\textbf{Measured and analytic domain leakage absolute error
      as a function of the iteration matrix spectral radius.}
    \textit{To test the behavior with respect to domain size, $n_i=50$
      (blue), $n_i=100$ (red), and $n_i=200$ (yellow) were used. The
      dashed lines represent the error using the Wigner rational
      approximation while the solid lines represent the error using
      the mean-chord approximation.}}
  \label{fig:measured_leakage}
\end{figure}

>>>>

  The domain decomposed behavior of the adjoint Neumann-Ulam method
  for solving linear operator equations is analyzed using the spectral
  properties of the linear operator. Relationships for the average
  length of the adjoint random walks and the fraction of histories
  leaking from a domain in the decomposition are made with respect to
  the Eigenvalues and other properties of the linear operator. In
  addition, an estimate of the effective optical thickness of a domain
  in the decomposition is presented based on the spectral analysis and
  diffusion theory. Using the leakage fraction estimate, Wigner
  rational approximation and the mean chord approximation are applied
  to estimate the leakage fraction. The one-speed, two-dimensional
  neutron diffusion equation is used as a model problem to test the
  spectral theory for symmetric operators. In general, the spectral
  theory and derived approximations show good agreement with measured
  computational results.

@@

\subsection{Domain Leakage}
Finally, we seek to measure the leakage from a domain in a domain
decomposed Monte Carlo calculation and assess the quality of our
analytic relation for the optical thickness of domain and the
associated leakage approximations. For this experiment, square grid
with $h=0.1$ and non-reentrant current boundary conditions. The grid
is then decomposed into 9 square domains, 3 in each cardinal
direction. For the cross sections, the absorption cross was varied
from 1 to 100 while the scattering cross section was set to zero to
create a purely absorbing environment. The optical thickness of these
domains will vary as a function of the absorption cross section if the
other parameters are fixed. To compute the optical thickness, along
with the spectral radius as given by Eq~(\ref{eq:iteration_radius}),
we also need the parameters $n_i$ and $n_s$ which respectively
describe the typical domain length and the average number of states
moved along that typical length per history transition. For our grid
above, the domains are varied in size with $50 \times 50$, $100 \times
100$, and $200 \times 200$ cells giving $n_i=50$, $n_i=100$, and
$n_i=200$ grid points or states along the typical length of the domain
respectively. Looking at the Laplacian stencil in
Figure~\ref{fig:stencil}, we see that all history transitions will
only move a single state in either the $i$ or $j$ directions due to
the symmetry of the problem. Furthermore, if we choose the $i$
direction, not all states we will transition to will move the history
in that direction. Therefore, we look to the definition of the
iteration matrix in Eq~(\ref{eq:iteration_stencil}) and the definition
of the adjoint probability matrix in Eq~(\ref{eq:adjoint_probability})
to estimate the $n_s$ parameter. For a particular transition starting
at state $(i,j)$, 6 of the 8 possible new states in the stencil move
the history in $i$ direction with relative coefficients of 4 for
moving in the $(\pm i,0)$ direction and of 1 for moving in the $(\pm
i,\pm j)$. These coefficients dictate the frequency those states are
visited relative to the others. For those 6 states we can visit along
the typical length their sum is 12 out of the total 20 for the
coefficients for all possible states with their ratio giving $n_i =
\frac{3}{5}$.

To compute the leakage fraction numerically, over the global domain
\sn{3}{5} histories were sampled from a uniform source of strength
unity. At the start of a stage of histories, the number of histories
starting in the domain was computed and as the stage progressed, the
number of histories that exited the domain was tallied with the ratio
of the two numbers providing a numerical measure for the leakage
fraction. Figure~\ref{fig:measured_length} gives the domain leakage
measurements for the domain in the center of the global grid as well
as the analytic result computed by
Eqs~(\ref{eq:wigner_domain_leakage}) and
(\ref{eq:mean_chord_domain_leakage}) as a function of the iteration
matrix spectral radius.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in,clip]{leakage_variation.png}
  \end{center}
  \caption{\textbf{Measured and analytic domain leakage as a function
      of the iteration matrix spectral radius.} \textit{To test the
      behavior with respect to domain size, $n_i=50$, $n_i=100$,and
      $n_i=200$ were used. The red data was computed numerically by a
      domain-decomposed adjoint Neumann-Ulam implementation, the black
      fine-dashed data was generated by
      Eq~(\ref{eq:mean_chord_domain_leakage}) using the mean-chord
      approximation, and the dashed-dotted black data was generated by
      Eq~(\ref{eq:wigner_domain_leakage}) using the Wigner rational
      approximation.}}
  \label{fig:measured_leakage}
\end{figure}
 Again, we note good qualitative agreement between the measured and
 analytic quantities but we begin to see the limits of these leakage
 approximations. To compare the quality of the two approximations, the
 absolute error between the computed leakage fraction and that
 generated by the Wigner rational and mean chord approximations is
 plotted in Figure~\ref{fig:leakage_error} for all domain sizes. From
 these error results, the mean chord approximation is shown to have a
 lower error for ill-conditioned systems as compared to the Wigner
 approximation while the Wigner approximation produces less error for
 more well-conditioned systems. In general, the mean chord
 approximation is a better choice to estimate the leakage fraction in
 a domain from the adjoint Neumann-Ulam method and except for a single
 data point with $n_i=50$, the mean chord approximation yielded
 leakage fractions within 0.05 of the measured results.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in,clip]{leakage_error.png}
  \end{center}
  \caption{\textbf{Measured and analytic domain leakage absolute error
      as a function of the iteration matrix spectral radius.}
    \textit{To test the behavior with respect to domain size, $n_i=50$
      (blue), $n_i=100$ (red), and $n_i=200$ (yellow) were used. The
      dashed lines represent the error using the Wigner rational
      approximation while the solid lines represent the error using
      the mean-chord approximation.}}
  \label{fig:leakage_error}
\end{figure}
